---
title: "wids datathon 2023"
output: github_document
---

___

### setup

```{r message = F, warning = F}
tictoc::tic()
library(tidyverse)
library(tidymodels)
library(tvthemes)
library(janitor)
library(lubridate)
library(skimr)
library(patchwork)
library(ggridges)
library(finetune)
library(vip)
options(scipen = 999)

theme_custom = theme_avatar() +
  theme(plot.title = element_text(hjust = 0.5),
        panel.grid.major = element_line(linewidth = 0.5, colour = "#D6D0C4"),
        panel.grid.minor = element_line(linewidth = 0.5, colour = "#D6D0C4"))
```

___

### data import

```{r}
train = read_csv("train_data.csv", col_types = cols()) |> clean_names()
# USING ONLY A TENTH OF THE DATA FOR RUNTIME PURPOSES
# DATATHON SUBMISSION WAS MADE USING ALL DATA
train = train |> slice_sample(prop = 0.1) # compute time
test = read_csv("test_data.csv", col_types = cols()) |> clean_names()
sample_sol = read_csv("sample_solution.csv", col_types = cols()) |> clean_names()
paste0("training dimensions: ", nrow(train), " rows x ", ncol(train), " columns")
paste0("testing dimensions: ", nrow(test), " rows x ", ncol(test), " columns")
```

___

### finding variable types

```{r}
paste0(ncol(select(train, where(is.numeric))), " numeric variables; ",
       ncol(select(train, where(is.character))), " character variables")
```

___

### reformatting `startdate` and `climateregions_climateregion` variables

```{r}
train = train |>
  mutate(startdate = parse_date(startdate, format = "%m/%d/%y"),
         year = year(startdate),
         month = month(startdate),
         day = day(startdate),
         climateregions_climateregion = as.factor(climateregions_climateregion)) |>
  rename(climate_region = climateregions_climateregion)

train |>
  select(startdate, year, month, day, climate_region) |>
  skim()
```

___

### counting climate regions

```{r}
train |>
  count(climate_region) |>
  mutate(pct = paste0(round(n / sum(n) * 100, 1), "%")) |>
  ggplot(aes(reorder(climate_region, -n), n)) +
  geom_col(aes(fill = climate_region)) +
  geom_text(aes(label = pct), size = 3, vjust = -0.5) +
  theme_avatar() +
  labs(x = "climate region", y = "count", title = "counts of climate regions") +
  theme_custom +
  theme(legend.position = "none")
```

___

### finding missing values

```{r}
data.frame(x = names(train), y = colSums(is.na(train))) |>
  filter(y > 0) |>
  ggplot(aes(reorder(x, y), y)) +
  geom_col(aes(fill = x)) +
  geom_text(aes(label = y), hjust = 1.25, size = 3) +
  labs(x = NULL, y = "number of NA values",
       title = "counts of missing values") +
  coord_flip() +
  theme_custom +
  theme(legend.position = "none")
```

___

### distribution of all missing values MAKE SURE TO REMOVE HEAD LINE

```{r}
vars_with_NAs = data.frame(col = names(train), n = colSums(is.na(train))) |>
  filter(n > 0) |>
  pull(col)

train |>
  select(index, all_of(vars_with_NAs)) |>
  pivot_longer(cols = -index, names_to = "var_name", values_to = "value") |>
  mutate(is_na = is.na(value)) |>
  ggplot(aes(var_name, index)) +
  geom_tile(aes(fill = is_na), alpha = 0.75) +
  scale_fill_manual(values = c("#71926B", "#BA5757"), labels = c("complete", "missing")) +
  coord_flip() +
  theme_custom +
  labs(x = NULL, y = "index", fill = NULL,
       title = "distribution of missing values - temporal pattern?") +
  theme(plot.title = element_text(hjust = 0.5),
        legend.position = "right")
```

___

### checking distributions of missing variables before imputation

```{r}
train |>
  select(all_of(vars_with_NAs)) |>
  pivot_longer(everything()) |>
  na.omit() |>
  ggplot(aes(value)) +
  geom_density(aes(fill = name), alpha = 0.4) +
  annotate("text", x = 95, y = 0.025, fontface = "italic", size = 3.5,
           label = "none of the variables look extremely\nskewed; we can impute with mean") +
  theme_custom +
  labs(title = "distributions of variables with missing values", fill = NULL) +
  theme(legend.position = "right")
```

___

### imputing missing data

```{r}
train = train |>
  mutate(across(.cols = all_of(vars_with_NAs), ~ replace_na(., mean(., na.rm = T))))

paste0("number of NA values in training data: ", sum(is.na(train)))
```

___

### visualization of target variable

```{r}
fig1 = train |>
  ggplot(aes(startdate, contest_tmp2m_14d_tmp2m)) +
  geom_line(col = "#718F70", alpha = 0.5) +
  geom_hline(yintercept = mean(train$contest_tmp2m_14d_tmp2m), col = "black") +
  annotate("text", x = as_date("2015-01-01"), y = mean(train$contest_tmp2m_14d_tmp2m) - 2,
           label = "avg. temp. (\u00B0C)", size = 3.5) +
  labs(x = NULL, y = "target variable temp. (\u00B0C)",
       title = paste0("average temperature over time: ", min(train$startdate),
                      " to ", max(train$startdate))) +
  theme_custom +
  coord_cartesian(expand = F)

fig2 = train |>
  ggplot(aes(startdate, y = 1)) +
  geom_tile(aes(fill = contest_tmp2m_14d_tmp2m)) +
  scale_fill_stepsn(colors = c("#789DBB", "#D6D0C4", "#D38585")) +
  theme_custom +
  labs(x = NULL, y = NULL) +
  theme(legend.position = "none",
        axis.text = element_blank()) +
  coord_cartesian(expand = F)

fig1 / fig2
```

___

### finding average monthly temperatures

```{r}
train |>
  mutate(month = month(month, label = T),
         year = factor(year)) |>
  group_by(year, month) |>
  summarise(mean_temp = mean(contest_tmp2m_14d_tmp2m),
            .groups = "drop") |>
  ggplot(aes(month, mean_temp)) +
  geom_line(aes(group = year, col = year), linewidth = 2, alpha = 0.5) +
  facet_wrap(vars(year), nrow = 3) +
  theme_custom +
  labs(x = "month", y = "average temperature (\u00B0C)",
       title = "average monthly temperature (\u00B0C) by year",
       subtitle = "note that 2015 is our only year of complete data") +
  theme(legend.position = "none",
        plot.subtitle = element_text(hjust = 0.5, vjust = 1.75, size = 9, face = "italic"))
```

___

### shapiro test for normality

```{r}
shapiro_sample = train |>
  pull(contest_tmp2m_14d_tmp2m) |>
  sample(5000) |>
  shapiro.test()

shapiro_sample
```

because of the p-value less than 0.05, we cannot assume that the population this data is sampled from is normally distributed.

___

### viewing distribution of target variable

```{r}
train |>
  ggplot(aes(contest_tmp2m_14d_tmp2m)) +
  geom_histogram(bins = 30, col = "black", fill = "#90B48F") +
  theme_custom +
  labs(x = "target variable: contest_tmp2m_14d_tmp2m",
       title = "histogram of target variable")
```

___

### distribution of target variable by year

```{r}
train |>
  mutate(year = factor(year)) |>
  ggplot(aes(contest_tmp2m_14d_tmp2m)) +
  geom_histogram(aes(fill = year), bins = 30, col = "black") +
  scale_fill_manual(values = c("#96B3CD", "#90B48F", "#BEA2C3")) +
  theme_custom +
  labs(x = "target variable: contest_tmp2m_14d_tmp2m",
       title = "histogram of target variable by year") +
  facet_wrap(vars(year), nrow = 3) +
  theme(legend.position = "none")
```

___

### distribution of target variable by region

```{r}
train |>
  mutate(year = factor(year)) |>
  ggplot(aes(contest_tmp2m_14d_tmp2m)) +
  geom_histogram(bins = 30, col = "black", fill = "#90B48F") +
  theme_custom +
  labs(x = "target variable: contest_tmp2m_14d_tmp2m",
       title = "histogram of target variable by year") +
  facet_wrap(vars(climate_region)) +
  theme(legend.position = "none")
```

___

### distribution of target variable by climate region

```{r message = F, warning = F}
train |>
  ggplot(aes(contest_tmp2m_14d_tmp2m, reorder(climate_region, contest_tmp2m_14d_tmp2m))) +
  geom_density_ridges(aes(fill = climate_region), col = "transparent", scale = 1) +
  theme_custom +
  labs(x = "target variable: contest_tmp2m_14d_tmp2m", y = "climate region",
       title = "distribution of target variable in different climate regions") +
  theme(legend.position = "none")
```

___

### distribution of target variable in climate regions by season

```{r message = F, warning = F}
train |>
  mutate(season = case_when(month %in% c(12, 1, 2) ~ "winter",
                            month %in% 3:5 ~ "spring",
                            month %in% 6:8 ~ "summer",
                            month %in% 9:11 ~ "fall")) |>
  ggplot(aes(contest_tmp2m_14d_tmp2m, reorder(climate_region, contest_tmp2m_14d_tmp2m))) +
  geom_density_ridges(aes(fill = climate_region), col = "transparent", scale = 1) +
  theme_custom +
  labs(x = "target variable: contest_tmp2m_14d_tmp2m", y = "climate region",
       title = "distribution of target variable in different climate regions by season") +
  theme(legend.position = "none") +
  facet_wrap(vars(season))
```

___

### visualizing variable correlations to target variable

```{r}
fig_df = train |>
  # sample_n(10000) |>
  select(where(is.numeric)) |>
  cor() |>
  data.frame() |>
  select(contest_tmp2m_14d_tmp2m)

fig_df |>
  mutate(var = rownames(fig_df)) |>
  ggplot(aes(reorder(var, contest_tmp2m_14d_tmp2m), contest_tmp2m_14d_tmp2m)) +
  geom_col(aes(fill = var)) +
  geom_hline(yintercept = -0.75, linetype = "dashed") +
  geom_hline(yintercept = 0.75, linetype = "dashed") +
  annotate("text", x = 10, y = 0.6, label = "threshold = 0.75", size = 3) +
  annotate("text", x = 240, y = -0.585, label = "threshold = -0.75", size = 3) +
  theme_custom +
  labs(x = "variable", y = "correlation with target variable",
       title = "variable correlations with target variable") +
  theme(legend.position = "none",
        axis.text.y = element_blank()) +
  coord_flip(ylim = c(-1, 1))
```

___

### time series visualization of target variable and relative humidity

```{r}
train |>
  ggplot() +
  geom_line(aes(startdate, contest_rhum_sig995_14d_rhum), alpha = 0.25, col = "#D18989") +
  geom_line(aes(startdate, contest_tmp2m_14d_tmp2m), alpha = 0.5, col = "#80A280") +
  scale_y_continuous(name = "target variable temperature (\u00B0C)",
                     sec.axis = sec_axis(~ . * 1, name = "relative humidity")) +
  theme_custom +
  labs(x = NULL, title = "time series of temperature and relative humidity") +
  theme(axis.title.y.left = element_text(color = "#658062"),
        axis.title.y.right = element_text(color = "#A56565"))
```

___

### time series visualization of temperature and precipitation

```{r}
train |>
  ggplot() +
  geom_line(aes(startdate, contest_precip_14d_precip / 10), alpha = 0.55, col = "#A5B7D4") +
  geom_line(aes(startdate, contest_tmp2m_14d_tmp2m), alpha = 0.5, col = "#80A280") +
  scale_y_continuous(name = "target variable temperature (\u00B0C)",
                     sec.axis = sec_axis(~ . * 1, name = "precipitation (divided by 10 for scale)")) +
  theme_custom +
  labs(x = NULL, title = "time series of temperature and precipitation") +
  theme(axis.title.y.left = element_text(color = "#658062"),
        axis.title.y.right = element_text(color = "#657794"))
```

___

### visualization of target variable by month

```{r message = F, warning = F}
ordered_months = c("2014-9", "2014-10", "2014-11", "2014-12",
                   "2015-1", "2015-2", "2015-3", "2015-4", "2015-5", "2015-6",
                   "2015-7", "2015-8", "2015-9", "2015-10", "2015-11", "2015-12",
                   "2016-1", "2016-2", "2016-3", "2016-4",
                   "2016-5", "2016-6", "2016-7", "2016-8")

train |>
  # slice_sample(prop = 0.1) |>
  mutate(month2 = paste0(year, "-", month),
         month2 = factor(month2, levels = rev(ordered_months))) |>
  ggplot(aes(contest_tmp2m_14d_tmp2m, month2)) +
  geom_density_ridges(aes(fill = factor(month)), col = "transparent", scale = 1) +
  theme_custom +
  labs(x = "target variable: contest_tmp2m_14d_tmp2m", y = "month",
       title = "distribution of target variable by month") +
  theme(legend.position = "none")
```

___

### visualizing relationship between precipitation and temperature

```{r}
cor_val = round(cor(train$contest_precip_14d_precip, train$contest_tmp2m_14d_tmp2m), 3)

train |>
  # slice_sample(prop = 0.1) |>
  ggplot(aes(contest_precip_14d_precip, contest_tmp2m_14d_tmp2m)) +
  geom_point(alpha = 0.25) +
  geom_smooth(formula = y ~ x, method = "lm", col = "springgreen4") +
  theme_custom +
  labs(x = "recorded precipitation", y = "recorded temperature",
       title = "scatterplot of recorded precipitation and temperature") +
  annotate("text", x = 250, y = -10, label = paste0("weak correlation;\ncoefficient = ", cor_val), size = 3.5, fontface = "italic")
```

___

### distribution of recorded precipitation

```{r message = F, warning = F}
unlogged = train |>
  ggplot(aes(contest_precip_14d_precip)) +
  geom_histogram(bins = 30, fill = "#90B48F", col = "black") +
  theme_custom +
  labs(x = "recorded precipitation", y = "count",
       title = "distribution of raw recorded precipitation v. log recorded precipitation")

logged = train |>
  mutate(log_precip = log(contest_precip_14d_precip)) |>
  ggplot(aes(log_precip)) +
  geom_histogram(bins = 30, fill = "#BEA2C3", col = "black") +
  theme_custom +
  labs(x = "log recorded precipitation", y = "count") +
  annotate("text", x = -2.5, y = 25000, label = "precipitation variable is\nhighly right-skewed", fontface = "italic", size = 3.5)

unlogged / logged
```

___

### climate regions

we have fifteen climate regions in the data; [here is what each value indicates](https://en.wikipedia.org/wiki/K%C3%B6ppen_climate_classification#:~:text=The%20K%C3%B6ppen%20climate%20classification%20scheme,indicates%20the%20level%20of%20heat.).

- BSh: hot semi-arid
- BSk: cold semi-arid
- BWh: hot desert
- BWk: cold desert
- Cfa: humid subtropical
- Cfb: temperature oceanic or subtropical highland
- Csa: hot summer Mediterranean
- Csb: warm summer Mediterranean
- Dfa: hot summer humid continental
- Dfb: warm summer humid continental
- Dfc: subarctic
- Dsb: Mediterranean-influenced warm summer humid continental
- Dsc: Mediterranean-influenced subarctic
- Dwa: monsoon-influenced hot summer humid continental
- Dwb: monsoon-influenced warm summer humid continental

___

### visualizing temperature and humidity by climate region

```{r}
train |>
  # slice_sample(prop = 0.1) |>
  ggplot(aes(contest_tmp2m_14d_tmp2m, contest_rhum_sig995_14d_rhum)) +
  geom_point(aes(col = climate_region)) +
  theme_custom +
  theme(legend.position = "top") +
  labs(x = "recorded temperature", y = "relative humidity",
       title = "relationship between temperature, humidity, and climate region")
```

___

### visualizing temperature and humidity, highlighting BSk climate region (most common in the data)

```{r message = F, warning = F}
train |>
  # slice_sample(prop = 0.1) |>
  ggplot(aes(contest_tmp2m_14d_tmp2m, contest_rhum_sig995_14d_rhum)) +
  geom_point() +
  geom_smooth(formula = y ~ x, method = "loess", se = F, col = "springgreen4", linewidth = 2) +
  gghighlight::gghighlight(climate_region == "BSk") +
  theme_custom +
  labs(x = "target variable temperature", y = "relative humidity",
       title = "scatterplot of temperature and humidity (BSk climate region highlighted)")
```

___

### visualizing elevation

```{r}
train |>
  ggplot(aes(elevation_elevation)) +
  geom_histogram(bins = 30, col = "black", fill = "#BBA3C1") +
  theme_custom +
  labs(x = "elevation", y = "count", title = "distribution of elevation")
```

___

### visualizing relationship between elevation and humidity

```{r}
elev_groups = train |>
  mutate(elev_group = cut_number(elevation_elevation, 5)) |>
  count(elev_group) |>
  pull(elev_group)

train |>
  mutate(elev_group = cut_number(elevation_elevation, 5),
         elev_group = case_when(elev_group == elev_groups[1] ~ "lowest",
                                elev_group == elev_groups[2] ~ "low",
                                elev_group == elev_groups[3] ~ "middle",
                                elev_group == elev_groups[4] ~ "high",
                                elev_group == elev_groups[5] ~ "highest"),
         elev_group = factor(elev_group, levels = c("lowest", "low", "middle", "high", "highest"))) |>
  ggplot(aes(elev_group, contest_rhum_sig995_14d_rhum)) +
  geom_boxplot(aes(fill = elev_group), outlier.alpha = 0.1, show.legend = F) +
  theme_custom +
  labs(x = "elevation group", y = "relative humidity",
       title = "boxplots of relative humidity by elevation group")
```

___

### visualizing relationship between relative humidity and measured precipitation

```{r}
train |>
  # slice_sample(prop = 0.1) |>
  ggplot(aes(contest_rhum_sig995_14d_rhum, contest_precip_14d_precip)) +
  geom_point() +
  geom_smooth(formula = y ~ x, method = "loess", se = F, col = "springgreen4") +
  theme_custom +
  labs(x = "relative humidity", y = "measured precipitation",
       title = "scatterplot of humidity and precipitation")
```

___

### visualizing temperature trends

```{r}
train |>
  group_by(startdate) |>
  summarise(mean_temp = mean(contest_tmp2m_14d_tmp2m)) |>
  mutate(temp_diff = mean_temp - lag(mean_temp),
         category = ifelse(temp_diff > 0, "warmer", "colder")) |>
  na.omit() |>
  ggplot(aes(startdate, temp_diff)) +
  geom_col(aes(fill = category)) +
  scale_fill_manual(values = c("#ADB8DF", "#CA8484")) +
  theme_custom +
  labs(x = "date", y = "difference temperature from past day", fill = "colder/warmer than day before",
       title = "time series of differences in daily average temperature") +
  theme(legend.position = "top")
```

___

### visualizing temperature by season

```{r}
train |>
  mutate(season = case_when(month %in% c(12, 1, 2) ~ "winter",
                            month %in% 3:5 ~ "spring",
                            month %in% 6:8 ~ "summer",
                            month %in% 9:11 ~ "fall"),
         season = factor(season, levels = c("summer", "fall", "winter", "spring"))) |>
  group_by(lat, lon, season) |>
  summarise(mean_temp = mean(contest_tmp2m_14d_tmp2m),
            .groups = "drop") |>
  ggplot(aes(lon, lat)) +
  geom_point(aes(col = mean_temp), size = 2) +
  scale_color_gradient(high = "#D58989", low = "#A6C2E2") +
  facet_wrap(vars(season)) +
  theme_custom +
  labs(x = "longitude", y = "latitude", title = "mapping temperature by season", col = "temperature")
```

___

### final variable adjustments before model pre-processing

```{r}
train = train |>
  select(-c(day, year)) |>
  mutate(month = factor(month, label = T))
```

___

### feature selection

```{r message = F, warning = F}
library(caret)
set.seed(123)
rPartMod = train(contest_tmp2m_14d_tmp2m ~ ., data = slice_sample(train, prop = 0.1), method = "rpart")
included_vars = varImp(rPartMod)[[1]] |> filter(Overall > 0) |> rownames()
included_vars
```

___

### creating final training set for modeling

```{r}
train_pre_selection = train
train = train |> select(contest_tmp2m_14d_tmp2m, index, lat, lon, startdate, all_of(included_vars))
glimpse(train)
```

___

### creating data split

```{r}
set.seed(123)
data_split = initial_split(train, prop = 0.75, strata = contest_tmp2m_14d_tmp2m)
train_data = training(data_split)
test_data = testing(data_split)

set.seed(234)
cv_folds = vfold_cv(train_data, v = 5, strata = contest_tmp2m_14d_tmp2m)
paste0("training data: ", nrow(train_data), " observations")
paste0("testing data: ", nrow(test_data), " observations")
```

___

### data preprocessing

```{r}
climate_rec = recipe(contest_tmp2m_14d_tmp2m ~ ., data = train_data) |>
  update_role(index, lat, lon, startdate, new_role = "ID") |>
  step_normalize(all_numeric(), -all_outcomes(), -index, -lat, -lon, -startdate)

summary(climate_rec)
```

___

### building model specification and workflow

```{r}
xgb_spec = boost_tree(trees = tune(), min_n = tune(), learn_rate = 0.25) |>
  set_engine("xgboost") |>
  set_mode("regression")

xgb_wf = workflow(climate_rec, xgb_spec)
xgb_wf
```

___

### tuning the model

```{r}
doParallel::registerDoParallel()
set.seed(345)

xgb_climate_rs = tune_grid(xgb_wf, cv_folds, grid = 10,
                           control = control_race(verbose_elim = T, pkgs = "stringr"))

xgb_climate_rs
```

___

### showing best models

```{r}
show_best(xgb_climate_rs, metric = "rmse")
```

___

### selecting best model

```{r}
xgb_last = xgb_wf |>
  finalize_workflow(select_best(xgb_climate_rs, "rmse")) |>
  last_fit(data_split)

xgb_last
```

___

### getting variable importance

```{r}
xgb_fit = extract_fit_parsnip(xgb_last)
vip(xgb_fit, geom = "point", num_features = 10) + theme_custom
```

___

### fitting final model and making predictions on test set

```{r}
best_tune = show_best(xgb_climate_rs, metric = "rmse") |>
  slice_min(mean)

n_trees = pull(best_tune, trees)
n_min_n = pull(best_tune, min_n)

data_recipe = recipe(contest_tmp2m_14d_tmp2m ~ ., data = train_data) |>
  update_role(index, lat, lon, startdate, new_role = "ID") |>
  step_normalize(all_numeric(), -all_outcomes(), -index, -lat, -lon, -startdate)

data_processed = prep(data_recipe) |>
  bake(train_data)

model = boost_tree(trees = n_trees, min_n = n_min_n, learn_rate = 0.25) |>
  set_engine("xgboost") |>
  set_mode("regression")

model_fit = fit(model, contest_tmp2m_14d_tmp2m ~ ., data = data_processed)
test = mutate(test, startdate = as_date(startdate, format = "%m/%d/%y"))

test_processed = prep(data_recipe) |>
  bake(test)

new_preds = predict(model_fit, new_data = test_processed)
final_df = data.frame(index = as.integer(test_processed$index), new_preds)
colnames(final_df) = c("index", "contest-tmp2m-14d__tmp2m")
write_csv(final_df, "submission.csv")

sample_n(final_df, 5)
```

### script runtime

```{r}
tictoc::toc()
```
































